PRISM protocol v0.3

### 2021-10-15 prism-backend/research/article.tex
Some rewriting. Added section on non-injective 'attack'.

In a nutshell, the proposed protocol is composed essentially of two phases. In
the first phase, the issuer shares a QR code with the holder. This QR code
contains the issuer's public key (or, maybe, its DID), a random value, and a
session identifier. This exchange takes place through a secure (i.e.,
confidential and authenticated) channel, such as physical exchange. Probably
immediately, although also possibly at a later point in time, the prospective
holder sends back her own public key (or, alternatively, her DID) and the
session identifier, all authenticated with a MAC, where we use the random value
in the QR code as secret key. The issuer uses the session identifier to locate
the corresponding session and, now that it received in an authenticated manner
the public key of the holder, can leverage hybrid encryption techniques to
establish an end-to-end encrypted session, e.g., using ECIES \cite{abr01}.
Ensuring end-to-end encryption and authentication is important at this point, as
all the information exchanged between issuer and holder is routed through the
connector: an entity or set of entities who ensure that asynchronous
communication can take place between issuer and holder. Thus, ECIES is crucial
to ensure the secrecy of the information, end-to-end. Since the holder needs to
have certainty that the data it receives comes from the issuer (and ECIES does
not authenticate the sender), we have the issuer digitally sign the data it
sends through the ECIES-secured channel. Note that we could involve the random
value shared via the QR code for this. But that would force us to re-execute the
first phase of the protocol even when the QR code is compromised after having
completed the first phase of the protocol. Digital signatures using the issuer's
DID prevent this.

@@

\begin{description}
\item[Code secrecy.] Codes exchanged via a QR code are always confidential.
  Note that this is trivial, as it is exchanged via a secure channel, and
  the holder never sends it back in the clear.
\item[Code injectivity.] Every QR code that is received by a holder, allegedly
  sent by an issuer, was sent before by that issuer. Again, this is trivial due
  to the secure channel used to exchange the QR code.
\item[Holder's public key authenticity.] Every message received by an issuer
  in step 3, containing a public key for key-exchange usage, that was
  allegedly sent by a holder, was indeed previously sent by that holder.
  This is ensured through the message authentication code (MAC) that
  leverages the random QR code exchanged via the secure channel, which is
  only known to the issuer and the holder.
\item[Holder's public key injectivity.] For every session in which an
  issuer receives in step 3 a message, allegedly sent by a holder in
  step 2, containing the holder's public key, it was that holder, in
  the same session, who sent the public key in step 2.  
\item[Credential secrecy.] This property has two variants: it ensures
  secrecy for credentials sent by an honest issuer to a holder that was honest
  at the moment of sending the credential, and is not compromised afterwards;
  and it also ensures secrecy for credentials received by an honest holder,
  allegedly from an honest issuer, if both holder and issuer were honest when
  the credential is received, and they are not corrupted afterwards. Altogether
  with the credential authenticity property shown next, it ensures that only
  (honest) holder and issuer learn the contents of a credential.
\item[Credential authenticity.] All messages containing a credential, received
  by a holder at step 5, where indeed sent by an issuer and addressed to that
  same holder, as long as both issuer and holder keys are not compromised.
\item[Credential non-injectivity.] All messages containing a credential, received
  by a holder at step 5, were sent by an issuer and addressed to that holder
  at step 4. Note, however, that both messages in steps 4 and 5 don't necessarily
  pertain to the  same ``session'' (in instantiations where parallel sessions
  may occur). However, this does not pose a risk, as in any case, the messaged
  is sent by the issuer, and received by the intended holder.
\end{description}

Concerning the technical details of the protocol, note that we are combining
signature and encryption (plus, the MAC-based authentication natively provided
by ECIES). This places us in the context put forward in \cite{davi01}, that
discusses some good practices when combining such primitives. In a nutshell,
we follow the sign-then-encrypt approach\footnote{What we call here ``sign-then-
  encrypt'' is called ``Sign \& Encrypt'' in \cite{davi01}. However, we find
  more explicit the naming convention of \cite{kraw01}, which is more consistent
  when referring to the order in which the primitives are used -- order that, in
  these cases, is crucial.}, combined with including the public
keys of both holder and issuer into the signed (and then encrypted) data.
This intertwines the inner layer (the signature) with the outer layer (the
encryption), ensuring that any alteration on any of them will be detected.
We note that there are other options that, a priori, are equally secure -- e.g.,
encrypt-then-sign, including also the sender's public key in the encrypted data
(but that would require separately signing a (hash of) the data, to ensure
non-repudiation). It is worth noting that ECIES follows the
encrypt-then-authenticate approach, which is proven to realize secure channels
in \cite{kraw01}, as long as the underlying encryption scheme is IND-CPA secure,
and the underlying MAC function is secure.

@@

>>>>

In a nutshell, the proposed protocol (see \figref{fig:e2econnector}) is composed
of two phases. In the first phase, the issuer shares a QR code with the holder.
This QR code contains the issuer's public key (or, maybe, its DID), a random
value ($code$, in \figref{fig:e2econnector}), and a session identifier ($token$,
in \figref{fig:e2econnector}). This exchange takes place through a secure (i.e.,
confidential and authenticated) channel, such as physical exchange. Probably
immediately, although also possibly at a later point in time, the prospective
holder sends back her own public key (or, alternatively, her DID) and the
session identifier, all authenticated with a MAC, where we use the random value
in the QR code as secret key. The issuer uses the session identifier to locate
the corresponding session and, now that it received in an authenticated manner
the public key of the holder, can leverage hybrid encryption techniques to
establish an end-to-end encrypted session, e.g., using ECIES \cite{abr01}.
Ensuring end-to-end encryption and authentication is important at this point, as
all the information exchanged between issuer and holder is routed through the
connector: an entity or set of entities who ensure that asynchronous
communication can take place between issuer and holder. Thus, ECIES is crucial
to ensure the secrecy of the information, end-to-end. Since the holder needs to
have certainty that the data it receives comes from the issuer (and ECIES does
not authenticate the sender), we have the issuer digitally sign the data it
sends through the ECIES-secured channel. Note that we could involve the random
value shared via the QR code for this. But that would force us to re-execute the
first phase of the protocol even when the QR code is compromised after having
completed the first phase of the protocol. Digital signatures using the issuer's
DID prevent this.

Note that we are combining signature and encryption (plus, the MAC-based
authentication natively provided by ECIES). This places us in the context put
forward in \cite{davi01}, that discusses some good practices when combining such
primitives. In a nutshell, we follow the sign-then-encrypt approach%
\footnote{What we call here ``sign-then-encrypt'' is called ``Sign \& Encrypt''
  in \cite{davi01}. However, we find
  more explicit the naming convention of \cite{kraw01}, which is more consistent
  when referring to the order in which the primitives are used -- order that, in
  these cases, is crucial.}, combined with including the public
keys of both holder and issuer into the signed (and then encrypted) data.
This intertwines the inner layer (the signature) with the outer layer (the
encryption), ensuring that any alteration on any of them will be detected.
We note that there are other options that, a priori, are equally secure -- e.g.,
encrypt-then-sign, including also the sender's public key in the encrypted data
(but that would require separately signing a (hash of) the data, to ensure
non-repudiation). Finally, it is worth noting that ECIES follows the
encrypt-then-authenticate approach, which is proven to realize secure channels
in \cite{kraw01}, as long as the underlying encryption scheme is IND-CPA secure,
and the underlying MAC function is secure -- but this is independent of the
sign-then-encrypt approach.

@@

\begin{description}
\item[Code secrecy.] Codes exchanged via a QR code are always confidential.
  Note that this is trivial, as it is exchanged via a secure channel, and
  the holder never sends it back in the clear.
\item[Code injectivity.] Every QR code that is received by a holder, allegedly
  sent by an issuer, was sent before by that issuer. Again, this is trivial due
  to the secure channel used to exchange the QR code.
\item[Holder's public key authenticity.] Every message received by an issuer
  in step 3, containing a public key for key-exchange usage, that was
  allegedly sent by a holder, was indeed previously sent by that holder.
  This is ensured through the message authentication code (MAC) that
  leverages the random QR code exchanged via the secure channel, which is
  only known to the issuer and the holder.
\item[Holder's public key injectivity.] For every session in which an
  issuer receives in step 3 a message, allegedly sent by a holder in
  step 2, containing the holder's public key, it was that holder, in
  the same session, who sent the public key in step 2.  
\item[Credential secrecy.] This property has two variants: it ensures
  secrecy for credentials sent by an honest issuer to a holder that was honest
  at the moment of sending the credential, and is not compromised afterwards;
  and it also ensures secrecy for credentials received by an honest holder,
  allegedly from an honest issuer, if both holder and issuer were honest when
  the credential is received, and they are not corrupted afterwards. Altogether
  with the credential authenticity property shown next, it ensures that only
  (honest) holder and issuer learn the contents of a credential.
\item[Credential authenticity.] All messages containing a credential, received
  by a holder at step 5, where indeed sent by an issuer and addressed to that
  same holder, as long as both issuer and holder keys are not compromised.
\item[Credential non-injectivity.] All messages containing a credential, received
  by a holder at step 5, were sent by an issuer and addressed to that holder
  at step 4. Note, however, that both messages in steps 4 and 5 don't necessarily
  pertain to the  same ``session'' (in instantiations where parallel sessions
  may occur). However, our current understanding is that this does not pose a
  risk, as in any case, the messaged is sent by the issuer, and received by the
  intended holder. For the sake of future reference, we give an illustrative
  flow of this ``non-injective'' execution, in \appref{app:non-inj-e2ee}.
\end{description}

@@

\appendix

\section{Example of Non-Injective Execution of Connector E2EE Protocol}
\label{app:non-inj-e2ee}

The diagram in \figref{fig:e2econnector-noninj} depicts an scenario in which a
non-injective execution between the issuer and the holder takes place.

First, we have to take into account that such execution requires a somehow
extended setting than what we considered in \secref{sssec:e2eeconnector}. Namely,
here we assume a (certainly realistic) scenario in which any prospective holder
has several computing devices -- e.g., a laptop and a smartphone. Similarly,
the issuer instantiates several concurrent services -- depicted as
\emph{Instance 1} and \emph{Instance 2} in \figref{fig:e2econnector-noninj}.
Still, the same user operates both computing devices, which share the same
private values; and all issuer instances have access to the same database with
holders' information, and the issuer's private keys.

\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.32]{figures/connector-noninjective.png}
  \caption{Sequence diagram of a sample non-injective execution of the E2EE
    connector protocol}
  \label{fig:e2econnector-noninj}%
\end{figure}

In the diagram, the holder requests a QR code once with each device -- for this,
we envision an extended setting in which the holder can request QR codes
securely (e.g., through signed email). The smartphone communicates with Instance
2 of the issuer, while the laptop communicates with Instance 1. Eventually, the
holder decides to complete the process from her laptop. The connector forwards
the request to Instance 1, which stores it in the database. Then, it might
happen that it is Instance 2 the one that takes care of finalizing the request:
it fetches the necessary data from the shared database, and sends the result to
the holder's phone (which is the computing device that communicated with
Instance 2). This lack of correspondence between who initiated a request, and
who responds it, is what is formally referred to as non-injectivity.
Theoretically, that might perfectly happen in our proposed protocol, as there is
nothing preventing such behaviour in a cryptographic manner, as our model shows.
However, two observations are relevant:

\begin{itemize}
\item The authenticity and secrecy properties still ensure that only the holder
  and issuer have access to the data that is sent by the issuer, and that it
  indeed originates from the issuer.
\item Such behaviour, if deemed unwanted, could be easily prevented through a
  proper session handling, at the implementation level.
\end{itemize}
